Learning Discrete-valued Bayesian Networks from Mixed Data

### 2015-11-11 BN_discretizer_paper.tex
started going through it

%%%%%%%%%%%%%%%%%%%%%%% file template.tex %%%%%%%%%%%%%%%%%%%%%%%%%
%
% This is a general template file for the LaTeX package SVJour3
% for Springer journals.          Springer Heidelberg 2010/09/16
%
% Copy it to a new file with a new name and use it as the basis
% for your article. Delete % signs as needed.
%
% This template includes a few options for different layouts and
% content for various journals. Please consult a previous issue of
% your journal as needed.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% First comes an example EPS file -- just ignore it and
% proceed on the \documentclass line
% your LaTeX will extract the file if required
\begin{filecontents*}{example.eps}
%!PS-Adobe-3.0 EPSF-3.0
%%BoundingBox: 19 19 221 221
%%CreationDate: Mon Sep 29 1997
%%Creator: programmed by hand (JK)
%%EndComments
gsave
newpath
  20 20 moveto
  20 220 lineto
  220 220 lineto
  220 20 lineto
closepath
2 setlinewidth
gsave
  .4 setgray fill
grestore
stroke
grestore
\end{filecontents*}
%
\RequirePackage{fix-cm}
%
%\documentclass{svjour3}                     % onecolumn (standard format)
%\documentclass[smallcondensed]{svjour3}     % onecolumn (ditto)
\documentclass[smallextended]{svjour3}       % onecolumn (second format)
%\documentclass[twocolumn]{svjour3}          % twocolumn
%
\smartqed  % flush right qed marks, e.g. at end of proof
%
\usepackage{amsmath}
\usepackage{bm}
\usepackage{graphicx} % Allows including images
\usepackage{epstopdf}
\usepackage{algorithm}
\usepackage[noend]{algpseudocode}
\usepackage[round]{natbib}
\usepackage{tikz}
\usetikzlibrary{bayesnet}
\setcitestyle{aysep={ }} %Author-Year SEParator
%
% \usepackage{mathptmx}      % use Times fonts if available on your TeX system
%
% insert here the call for the packages your document requires
%\usepackage{latexsym}
% etc.
%
% please place your own definitions here and don't use \def but
% \newcommand{}{}
%
% Insert the name of "your journal" with
% \journalname{myjournal}
%
\begin{document}

@@

\begin{abstract}
Insert your abstract here. Include keywords, PACS and mathematical
subject classification numbers as needed.
\keywords{Discretization \and Bayesian Network \and Continuous Variable}
% \PACS{PACS code1 \and PACS code2 \and more}
% \subclass{MSC code1 \and MSC code2 \and more}
\end{abstract}

\section{Introduction}
\label{intro}
Bayesian networks \citep{Pearl_1988, PGM_2009}  are an increasingly popular method for modeling uncertainty and causality in science and engineering. They provide an efficient factorization of the joint probability distribution over a set of random variables. Bayesian networks first emerged from artificial intelligence research and have been applied to a wide variety of problems, ranging from decision-making systems \citep{DMU_2015} to medical diagnoses. In most cases, we assume that all random variables in Bayesian networks are discrete, since many algorithms on Bayesian networks are unable to deal with continuous variables efficiently. However, the assumption that all variables are discrete is often too restrictive. For example, in the decision-making system of autonomous cars, it is imperative to deal with continuous variables such as position and velocity.

There are two methods around this assumption. The first one is to model conditional probability density of each continuous variable by specific families of parametric distributions, then redesign algorithms on Bayesian networks based on these parameters. One successful example is belief propagation in Gaussian graphical models \citep{Weiss_2011}. Nevertheless, for other shapes of particles \citep{Ihler_2009} or non-linear functions, these redesigned algorithms are computationally expensive and do not perform well.

The second method around continuous variables is to discretize them. Discretization that learns from data has been developed well and discussed in the fields of machine learning and statistics for many years \citep{Dougherty_1995, Kerber_1992, Holte_1993, Fayyad_1993} . Most of these discretization methods are designed for classification problems. They search the best discretization policy of a continuous attribute by considering its interation with the class variable of interest. However, these discretization methods do not apply to continuous variables in Bayesian networks. In Bayesian networks, interations and dependencies between variables are determined by graph structure. Therefore, if a method discretizes continuous variables according to graph structure, instead of assigning one variable as a class variable, then it would be more appropriate. There is some reseach on discretizing continuous variables in naive Bayesian networks and tree augmented networks \citep{Fried_naive}, but only a few discretization methods on general Bayesian netowrk have been proposed \citep{Friedman_1996, Kozlov_1997, Monti_1998, Steck_2007}.

The discretization technique proposed by \citet{Friedman_1996} is the most well-known one among these few methods. It is based on minimum description length principle (MDL): optimal discretization policy minimizes the description length of Bayesian network and data information. If there is only one continuous variable in a Bayesian network and other variables are discrete, this MDL discretization method takes running time $O(N^3 + n_c {v_{max}}^{(n_c^p)_{max}} \cdot N^2 + {v_{max}}^{n_p} \cdot N^2)$, where $N$ is the number of data instances for learning the discretization, $n_p$ and $n_c$ are the numbers of parent and children variables for the continuous variable, $v_{max}$ is the largest cardinality over all variables in the Markov blanket, and  ${(n_c^p)_{max}}$ is the largest number of parent variables of the continuous variable's children. If there are multiple continuous variables in the network, the method iterates over all continuous variables until convergence. In our test on real-world data, the iterations converge in a few cycles. However, the number of intervals after discretization is usually too small.

In this paper, we propose a new discretization method for continuous variables in Bayesian networks by learning from mixed data. This method is a generalized version of  \citet{Boulle_2006} and \citet{Lustgarten_2011}, which are both discretization methods of a continuous attribute with one class variable. We begin our method with the assumption that only one variable in a given Bayesian network is continuous and other variables are discrete. Furthermore we assume that the network structure is known in advance. Under this situation, we look for the most probable discretization policy $M$ given the data $D$ on other discrete variables. That is to say, the desired policy is $\textrm{arg}_M P(M \mid D)$. With Bayes' rule, $P(M \mid D) $ can be rewritten as $P(M) \cdot P(D \mid M) / P(D)$ , which is proportional to $P(M)\cdot P(D \mid M)$. Usually, $P(D \mid M)$ increases as the number of discretized intervals increases, since more intervals can provide more accuracy. $P(M)$, on the other hand, is designed to decrease as the number of intervals rises. As a result, we can automatically determine the number of intervals after discretization by maximizing  $P(M)\cdot P(D \mid M)$. In addition, the $P(D \mid M)$ can be factorized according to Bayesian network structure. With the proposed priors, we are able to restrict the number of discretized intervals so that it will not exceed the largest cardinality of variables in the Markov blanket by too much. This is important, since for most algorithms on Bayesian networks, their running times exponentially depend on the cardinality of variables. Another advantage of our method is the running time for learning. If there is only one continuous variable, the running time is $O(n_c  {v_{max}} \cdot N^2 + {v_{max}}^{n_p} \cdot N^2)$, which is significantly shorter than the running time of MDL principle discretization \citep{Friedman_1996}. With the approximation proposed in this paper, we can further reduce running time to $O({v_{max}}(n_p+n_c)N^2)$.

For a Bayesian network with multiple continuous variables, we apply the discretization method discussed above iteratively. In the beginning, we prediscretize each continuous variable by equal-width method. The number of intervals for equal-width discretization is equal to largest cardinality of discrete variables. Then we iterate the one-variable discretization on each continuous variable in the following order: from the variable with highest topological order (leaves) to the variable with lowest topological order (root). Each time we finish a discretization procedure on one variable, we store the discretization result for later iterations. Experiments on real-world data show that with the same iteration order, our discretization method provides better discretization results than MDL principle method in terms of likelihood. The MDL principle method is easily stuck at local minima and also discretizes variables into too few intervals.

Finally, we can combine our new discretization method with the K2 structure learning alorithm \citep{K2}. We first prediscretize all continuous variables before the K2 learning, since the K2 algorithm requires all variables to be discrete. Each time an edge is added in the K2 algorithm procedure, we rediscretize the continuous variables based on the updated Bayesian network and store the result for the next K2 algorithm iteration. By this principle, we are able to learn a discrete-valued Bayesian network from mixed data.

@@

% Non-BibTeX users please use
%\begin{thebibliography}{}
\bibliographystyle{plainnat}
\bibliography{my_bib}

>>>>

%%%%%%%%%%%%%%%%%%%%%%% file template.tex %%%%%%%%%%%%%%%%%%%%%%%%%
%
% This is a general template file for the LaTeX package SVJour3
% for Springer journals.          Springer Heidelberg 2010/09/16
%
% Copy it to a new file with a new name and use it as the basis
% for your article. Delete % signs as needed.
%
% This template includes a few options for different layouts and
% content for various journals. Please consult a previous issue of
% your journal as needed.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% First comes an example EPS file -- just ignore it and
% proceed on the \documentclass line
% your LaTeX will extract the file if required
\begin{filecontents*}{example.eps}
%!PS-Adobe-3.0 EPSF-3.0
%%BoundingBox: 19 19 221 221
%%CreationDate: Mon Sep 29 1997
%%Creator: programmed by hand (JK)
%%EndComments
gsave
newpath
  20 20 moveto
  20 220 lineto
  220 220 lineto
  220 20 lineto
closepath
2 setlinewidth
gsave
  .4 setgray fill
grestore
stroke
grestore
\end{filecontents*}
%
\RequirePackage{fix-cm}
%
%\documentclass{svjour3}                     % onecolumn (standard format)
%\documentclass[smallcondensed]{svjour3}     % onecolumn (ditto)
\documentclass[smallextended]{svjour3}       % onecolumn (second format)
%\documentclass[twocolumn]{svjour3}          % twocolumn
%
\smartqed  % flush right qed marks, e.g. at end of proof
%
\usepackage{amsmath}
\usepackage{bm}
\usepackage{graphicx} % Allows including images
\usepackage{epstopdf}
\usepackage{algorithm}
\usepackage[noend]{algpseudocode}
\usepackage[round]{natbib}
\usepackage{tikz}
\usetikzlibrary{bayesnet}
\setcitestyle{aysep={ }} %Author-Year Separator
%
% \usepackage{mathptmx}      % use Times fonts if available on your TeX system
%
% insert here the call for the packages your document requires
%\usepackage{latexsym}
% etc.
%
% please place your own definitions here and don't use \def but
% \newcommand{}{}
%
% Insert the name of "your journal" with
% \journalname{myjournal}
%

\newcommand{\todo}[1]{\textcolor{magenta}{#1}}
\newcommand{\tim}[1]{\textit{\textcolor{blue}{#1}}}

\begin{document}

@@

\begin{abstract}
\todo{
Insert your abstract here. Include keywords, PACS and mathematical
subject classification numbers as needed.
}
\keywords{Discretization \and Bayesian Network \and Continuous Variable}
% \PACS{PACS code1 \and PACS code2 \and more}
% \subclass{MSC code1 \and MSC code2 \and more}
\end{abstract}

\section{Introduction}
\label{intro}
Bayesian networks \citep{Pearl_1988, PGM_2009} are an increasingly popular method for modeling uncertainty and causality in science and engineering. They provide an efficient factorization of the joint probability distribution over a set of random variables. Bayesian networks first emerged from artificial intelligence research and have been applied to a wide variety of problems, ranging from decision-making systems \citep{DMU_2015} to medical diagnoses~\todo{[cite]}. In most cases, we assume that all random variables in Bayesian networks are discrete, since many Bayesian network learning algorithms are unable to deal with continuous variables efficiently. However, many applications
require the use of continuous variables, such as position and velocity in dynamic systems.

There are two common approaches to extending Bayesian networks to continuous variables. The first is to model the conditional probability density of each continuous variable using specific families of parametric distributions, and then to redesign Bayesian network learning algorithms based on their parameters. One successful example is belief propagation in Gaussian graphical models \citep{Weiss_2011}. For other particle shapes \citep{Ihler_2009} and non-linear functions \tim{cite if you can}, these redesigned algorithms are computationally expensive and do not perform well.
\tim{Not sure what `particle shapes' are. Do you mean other parametric distributions?}

The second approach is discretization. Automated discretization methods have been studied in machine learning and statistics for many years \citep{Dougherty_1995, Kerber_1992, Holte_1993, Fayyad_1993}. Most discretization methods were designed for classification problems. They search for the best discretization policy\tim{maybe `scheme' is a better word than `policy', which in machine learning means something else.} of a continuous attribute by considering its interation with the target class variable. These discretization methods do not apply to continuous variables in Bayesian networks, where interations and dependencies between variables are more complicated. Discretization of a continuous variable in a Bayesian network requires considering the impact on the entire network. There is prior work on discretizing continuous variables in naive Bayesian networks and tree-augmented networks \citep{Fried_naive}, but only a few discretization methods for general Bayesian networks have been proposed \citep{Friedman_1996, Kozlov_1997, Monti_1998, Steck_2007}.

The discretization technique proposed by \citet{Friedman_1996} is the most well-known of these methods. It is based on the minimum description length principle (MDL): the optimal discretization policy minimizes the description length of the Bayesian network and discretization scheme. The MDL algorithm has a $O(N^3 + n_c {v_{max}}^{(n_c^p)_{max}} \cdot N^2 + {v_{max}}^{n_p} \cdot N^2)$ run time on a Bayesian network with a single continuous variable and multiple discrete variables, where $N$ is the number of data samples from the joint distribution, $n_p$ and $n_c$ are the numbers of parent and children variables for the continuous variable, $v_{max}$ is the largest cardinality over all variables in the Markov blanket, and  ${(n_c^p)_{max}}$ is the largest number of parent variables of the continuous variable's children. The MDL method iterates over all continuous variables until convergence in the case of multiple continuous variables, converging in a few cycles according to our tests of real-world data. Unfortunately, MDL tends to produce discrete variables with low cardinality.

In this paper, we propose a new Bayesian discretization method for continuous variables in Bayesian networks. This method extends the Bayesian single-variable discretization methods from \citet{Boulle_2006} and \citet{Lustgarten_2011}.

% We begin our method with the assumption that only one variable in a given Bayesian network is continuous and other variables are discrete. Furthermore we assume that the network structure is known in advance. Under this situation, we look for the most probable discretization policy $M$ given the data $D$ on other discrete variables. That is to say, the desired policy is $\textrm{arg}_M P(M \mid D)$. With Bayes' rule, $P(M \mid D) $ can be rewritten as $P(M) \cdot P(D \mid M) / P(D)$ , which is proportional to $P(M)\cdot P(D \mid M)$. Usually, $P(D \mid M)$ increases as the number of discretized intervals increases, since more intervals can provide more accuracy. $P(M)$, on the other hand, is designed to decrease as the number of intervals rises. As a result, we can automatically determine the number of intervals after discretization by maximizing  $P(M)\cdot P(D \mid M)$. In addition, the $P(D \mid M)$ can be factorized according to Bayesian network structure. With the proposed priors, we are able to restrict the number of discretized intervals so that it will not exceed the largest cardinality of variables in the Markov blanket by too much. This is important, since for most algorithms on Bayesian networks, their running times exponentially depend on the cardinality of variables. Another advantage of our method is the running time for learning. If there is only one continuous variable, the running time is $O(n_c  {v_{max}} \cdot N^2 + {v_{max}}^{n_p} \cdot N^2)$, which is significantly shorter than the running time of MDL principle discretization \citep{Friedman_1996}. With the approximation proposed in this paper, we can further reduce running time to $O({v_{max}}(n_p+n_c)N^2)$.

% For a Bayesian network with multiple continuous variables, we apply the discretization method discussed above iteratively. In the beginning, we prediscretize each continuous variable by equal-width method. The number of intervals for equal-width discretization is equal to largest cardinality of discrete variables. Then we iterate the one-variable discretization on each continuous variable in the following order: from the variable with highest topological order (leaves) to the variable with lowest topological order (root). Each time we finish a discretization procedure on one variable, we store the discretization result for later iterations. Experiments on real-world data show that with the same iteration order, our discretization method provides better discretization results than MDL principle method in terms of likelihood. The MDL principle method is easily stuck at local minima and also discretizes variables into too few intervals.

% Finally, we can combine our new discretization method with the K2 structure learning alorithm \citep{K2}. We first prediscretize all continuous variables before the K2 learning, since the K2 algorithm requires all variables to be discrete. Each time an edge is added in the K2 algorithm procedure, we rediscretize the continuous variables based on the updated Bayesian network and store the result for the next K2 algorithm iteration. By this principle, we are able to learn a discrete-valued Bayesian network from mixed data.

@@

% Non-BibTeX users please use
%\begin{thebibliography}{}
\bibliographystyle{spbasic}
\bibliography{my_bib}
